PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2023-02-24 reviewer_response.tex

\reviewer{R3}{In the pseudo-energy discussion a further parameter is mentioned: "In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.". The value chosen for this parameter should be explicitly stated along with a rationale for the choice. It is also not clear if any of the splitting pancontigs are less than this value if the entire merger is dropped or just massaged to not have the too small new pancontig.}
% \Liam{this relates to R2's point above about all sequence ending up in pancontigs so I think is just clarification.}\\
% \Marco{I agree. We set a minimal cutoff of 100bp. This is stated later in line 106. Maybe we could expand the sentence in line 99 and give the value and few more words on the meaning/effect. This controls in some sense the ``corse graining'' of the graph: what is the size of an event that we want to encode as structural variation in the graph, rather than as an indel in the alignment. In the algorithm this minimal cutoff has different effects depending on the context. When aligning, if the match lenght is less than this threshold the merge is refused. For long alignments, if the CIGAR string contains indels that are longer than this threshold, these are saved as separate blocks and not as indels in the block. The same goes for the flanking sequences, if merging two long blocks leaves overhangs that are less than this threshold size then these overhangs are saved as part of the merged block and not as separate small blocks. This might be explained in the SI?}\\
\responsedraft{
      The minimal size parameter $\Lthr$ is set to 100 bp in PanGraph. This parameter controls the minimal size of blocks. In general, this parameter sets a scale separation between structural variation that is encoded in the graph structure or in the pancontig alignment. For instance, an insertion could appear either as a separate pancontig, or as an insertion in the pancontig alignment, depending on its size. In practice this parameter is used in three ways:
      \begin{itemize}
            \item In addition to the pseudo-energy costraint, merges are refused if they have alignment length shorter than $\Lthr$.
            \item During merging, if the alignment contains long ($> \Lthr$) indels, these will cause a split in the merged pancontigs and will be treated as separate pancontigs.
            \item If a merger of two pancontigs would produce flanking overhangs shorter than $\Lthr$, these are not saved as separate pancontigs but are included in the alignment as flanking indels.
      \end{itemize}
      As pointed out by the reviewer this parameter was not properly discussed in the text. We now extended its definition in IB in the main text, and give a more extended discussion in SI section I.\\
      % TODO: missing: motivation for this choice in the text.
      % TODO: Homogenization in the main text with discussion on precision. Extend.
}

### 2023-02-24 main.tex
save progress

\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\Lcore}{L_\text{core}}
\newcommand{\Lsoftcore}{L_\text{95\%}}
\newcommand{\Lpang}{L_\text{pang}}
\newcommand{\Lgen}{L_\text{gen}}
\newcommand{\Lthr}{L_{\min}}
\newcommand{\dcore}{\langle d_\text{core} \rangle}
\newcommand{\modified}[1]{{\color{red}#1}}
\newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}

@@

%\section{Introduction}
During evolution, microbial genomes change by both local mutations and large-scale alterations \cite{arnold2021horizontal}.
Local mutations only change a few nucleotides by substitution, insertion or deletion.
Conversely, large-scale alterations reorganize the sequence, and involve either the homologous recombination of large segments, gene loss or gain, inversions, or mobilisation of genetic elements.
The accumulation of such changes over time complicates comparative genomic analyses of present day isolates.
Homologous recombination is rapid enough that most genes in many bacterial core genomes have distinct phylogenies \cite{sakoparnig2021whole} and even closely related genomes differ dramatically in gene content.
% comment: rewritten to use consistent terminology (I think 'mutation' and 'polymorphism' for most people make them think of local changes only)
%Scalable inference techniques of this complicated relatedness structure would, in part, spur the advance of critical epidemiological surveillance tools, analogous to NextStrain, for bacterial pathogens.

@@

\section{Algorithms and implementation}
\emph{PanGraph} transforms an arbitrary set of genomes into a \emph{graph} that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms.
The graph is composed of \emph{pancontigs}, which represent linear multiple-sequence alignments of homologous sequence found within one or more input genomes.
\emph{Pancontigs} are connected by an edge if they are \oldnew{syntenic on}{adjacent in} at least one input sequence; individual sequences are then recapitulated by contiguous \emph{paths} through the graph.
Pancontigs are directed, meaning their orientation in a path can be either 5' to 3' or vice versa.

To construct a pangraph, the algorithm needs to find homologous sequence within and among all input genomes.
\emph{PanGraph's} overarching strategy is to approximate multiple-genome alignment by iterative pairwise alignment of graphs of subsets of sequences, in the spirit of progressive alignment tools \cite{feng1987progressive,darling2010progressivemauve,armstrong2020progressive}.
Pairwise graph alignment is performed by an all-to-all alignment of the \emph{pancontigs} consensus sequences between both graphs, and the order of pairwise alignments is determined by a guide tree.

@@

\paragraph*{Merging of homologous sequence.}
If the above alignment step detected homologous stretches between \oldnew{}{the consensus sequences of} two pancontigs, these pancontigs, or parts of them, can be merged.
It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.
We rank each alignment between two pancontigs according to the pseudo-energy
\begin{equation}\label{eq:pseudo-energy}
    E = -\ell + \alpha N_c + \beta N_m
\end{equation}
where $\ell$, $N_c$, and $N_m$ denote the alignment length, number of \emph{pancontigs} created by the merger, and number of polymorphisms per genome in the newly created {pancontig} respectively.
See Fig \ref{fig:visualization}B for a graphical definition for each term.
Additionally, $\alpha$ and $\beta$ are hyperparameters of the algorithm.
Only mergers whose alignment has negative pseudo-energy are performed.
The parameter $\beta$ controls the maximal sequence divergence between the two merger-candidates\oldnew{ and no mergers will be performed if the Hamming distance exceeds $1/\beta$.}{. The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.}
Importantly, this divergence threshold is applied to the comparison of the consensus sequences of the merger candidates.
\oldnew{}{The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.}
The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the \oldnew{resulting}{} pan-contigs resulting from the merger.
The merger is always performed if \oldnew{the merger}{it} doesn't increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
\oldnew{In addition, there is parameter that controls the minimal size of contigs produced by the algorithm.}{In addition to the fragmentation parameter $\alpha$, there is a further parameter $\Lthr$ that controls the minimal length of pancontigs. Structural variation at a scale smaller than this threshold (i.e. short indels) is stored in pancontig alignments, rather than in the graph structure, and no merges are performed if they have length shorter than this threshold. We set a default value of 100 bp for this parameter.}

@@

At the graph level, the merger of two \emph{pancontigs} defines a new \emph{pancontig}, connected on both sides by edges to the neighboring \emph{pancontigs} of both inputs, and thus locally collapses the two graphs under consideration.
At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the reference of one onto the other; the merger of two \emph{pancontigs} requires the application of the map onto the underlying multiple-sequence alignment.
Once both sets of sequences are placed onto a common coordinate system, the resultant consensus sequence, and thus polymorphisms, are recomputed.
This procedure can be viewed as an approximate multiple sequence alignment algorithm that aligns homologous parts without further investigating insertions in the two sub-alignments.

The above procedure is repeated until no alignments with negative energy remain.
Upon completion, transitive edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent \emph{pancontigs}. \oldnew{}{More details on the merging procedure and the effect of parameters can be found in SI section I.}

@@

\begin{figure}[tb]
    \includegraphics[width=.4\textwidth]{figs/misplaced_fraction_vs_divergence.pdf}
    \caption{{\bf Accuracy against synthetic data}.
        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels \textit{minimap2} with \textit{asm10} or \textit{asm20} option, and \textit{mmseqs2}. In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp. The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of \oldnew{}{$\Lthr=$}100bp, as a function of average pairwise sequence divergence. Line and shaded area represents mean and standard deviation over 25 repetitions. \textit{mmseqs2} maintains accuracy at higher divergence, at the cost of higher computational time.
    }
    \label{fig:toy-accuracy}
\end{figure}

@@

\begin{table}[h]
    \setlength{\tabcolsep}{6pt}
    \begin{tabular}{l c c c c c c}
        \hline\hline
        Species                  & N.  & $\avg{\Lgen}$ & $\Lpang$ & $\Lcore$ & $\Lsoftcore$ & $\dcore$ \\
                                 &     & [Mbp]         & [Mbp]    & [Mbp]    & [Mbp]        &          \\
        \hline
        \textit{E.~coli}         & 307 & 5.0           & 17.7     & 0.7      & 2.0          & 1.6\%    \\
        \textit{K.~pneumoniae}   & 109 & 5.3           & 13.0     & 2.3      & 3.7          & 0.8\%    \\
        \textit{H.~pylori}       & 85  & 1.6           & 1.9      & 0.6      & 1.0          & 4.2\%    \\
        \textit{M.~tuberculosis} & 51  & 4.4           & 4.1      & 2.4      & 3.4          & 0.03\%   \\
        \textit{P.~marinus}      & 10  & 1.8           & 3.3      & 0.7      & 0.7          & 26.9\%   \\
        \hline
    \end{tabular}
    \caption{{\bf Dataset properties}. Columns represent:
        number of isolates (N.),
        average chromosome length ($\avg{\Lgen}$),
        total pangenome length ($\Lpang$),
        total core genome length ($\Lcore$),
        \oldnew{}{total soft-core genome length ($\Lsoftcore$, corresponding to genes shared by at least 95\% of the isolates),}
        and average pairwise divergence of core genes ($\dcore$).
    }
    \label{table:panx-dataset}
\end{table}

@@

For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in \oldnew{blocks}{pancontigs} spanning several kbps (see SI Fig.~4).
For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf. SI Section IIIC and SI Fig.~4).
Overall this benchmark showcases the capabilities and limits of \textit{PanGraph}, and demonstrates the value of adapting the pseudo-energy parameter and choice of alignment kernel to the expected diversity of the dataset considered.

\begin{figure*}[htb]
    \includegraphics[width=\textwidth]{figs/panx_benchmark.pdf}
    \caption{{\bf Benchmark on real data}.
        We build pangenome graphs from fully-assembled chromosomes from 5 different bacterial species. For each species we build graphs with three different alignment kernel options (\textit{minimap2} with \textit{asm10} or \textit{asm20} option and \textit{mmseqs2}) and two different settings for the pseudo-energy parameters $\alpha, \beta$ (standard or null values).
        \textbf{A}: \textit{PanGraph} wall-time when run in parallel on 8 cores.
        \textbf{B}: fraction of core pangenome in the pangenome graph.
        \textbf{C}: sequence compression, defined as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph. \oldnew{}{Since maximal compression depends on the number of isolates $n$ in the pangenome graph, we mark for reference the value of $1/n$ for each dataset.}
    }
    \label{fig:panx-benchmark}
\end{figure*}

@@

We then explored how the properties of the pangraph scale with the dataset size. To do so, we build pangenome graphs from randomized subsets of isolates of increasing size from the \textit{E.~coli} dataset. Results are reported in Fig.~\ref{fig:panx-size}.\\
The number and size of pancontigs scale sub-linearly with the number of isolates, with more than 50\% of the pangenome being included in the 10\% longest pancontigs. \oldnew{, and core pancontigs having higher-than-average size}{Core pancontigs have higher-than-average length. This is expected given their strong synteny and the fact that they present less structural variation across genomes compared to accessory genes.} Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added. The size of the core genome does not decrease significantly with the addition of new strains, while the total pangenome size remains orders of magnitude smaller than the total size of all genomes included in the graph.

@@

In addition to being a useful operation, marginalization allows us to quantify the consistency and robustness of pangraph construction on real data, where the ground truth is unknown.
To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{K.~pneumoniae} dataset. We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair (i.e. marginalized graph in Fig.~\ref{fig:marginalization}A top), and the graph built directly from the pair (i.e. pairwise graph in Fig.~\ref{fig:marginalization}A top).
In order to compare both graphs, we compute the partition they generated on the genome of the isolates they include. Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate (cf. Fig.~\ref{fig:marginalization}A bottom). We can classify the segments in the intersection of the two partitions in different categories, depending on whether the two partitions agree on whether the segment is shared or private. Moreover segments on which the two partitions agree are further sub-divided depending on whether they are private or shared on both partitions.\\
The compatibility between the two graphs requires segments on which the two partitions disagree to be few and short. Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} \oldnew{}{set by the value of $\Lthr$} (cf. Fig.~\ref{fig:marginalization}B-C). Conversely, segments on which the partitions agree have average size of several kbps. The fact that the two partitions almost completely agree cannot be explained simply by the fact that most of the genome is shared, since segments that are shared on both graphs cover on average only 85\% of the sequence. We confirmed these results using the fraction of shared k-mers, using an approach inspired by PopPUNK \cite{lees2019fast}. Namely, we approximate the fraction of homologous sequence between any two pairs using the fraction of shared k-mers ($k=21$) divided by $(1-d)^{k}$, where $d$ is the average pairwise divergence on core genes for the pair considered. This divisor corrects for k-mers on homologous sequence that are not shared due to mutations. The resulting distribution of shared sequence is in very good agreement with the fraction of segments that are shared on both graphs (cf. Fig.~\ref{fig:marginalization}B), suggesting that homologous sequences are correctly merged on the complete, marginalized and pairwise graphs.\\
We performed the same test on species from the other datasets, obtaining similar results on all species whose divergence is compatible with \textit{PanGraph} capabilities (cf. SI Section IV and SI Fig.~5).

\begin{figure*}[htb]
    \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
    \caption{{\bf Test of graph marginalization}.
        \textbf{A}: we built a pangenome graph from 50 randomly chosen strains from the \textit{K.~pneumoniae} dataset. We then randomly picked 50 pairs of strains. For each pair we compared the pangenome graph obtained by marginalizing the complete graph on the pair of strain, and the one obtained by building a new graph for the pair (top). The comparison is done by considering that each graph partitions a genome in shared and private segments. By combining the partitions generated by the marginalized and pairwise graphs we categorize segments in three categories, depending on whether the two partitions agree or not, and if they agree depending on whether segments are shared or private. All graphs were built using \textit{minimap2} alignment kernel with \textit{asm20} option and default value for the energy parameters.
        \textbf{B}: distribution of the average fraction of the genome covered by segments of each category, over the 50 pairs considered (2 entries per pair). The last line represents the distribution of shared sequence, approximated using the fraction of shared k-mers corrected using sequence divergence as described in the main text. \oldnew{}{Next to each distribution we report its mean and standard deviation.}
        \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair). \oldnew{}{Mean and standard deviation are reported.}
    }
    \label{fig:marginalization}
\end{figure*}

>>>>

\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\Lcore}{L_\text{core}}
\newcommand{\Lsoftcore}{L_\text{95\%}}
\newcommand{\Lpang}{L_\text{pang}}
\newcommand{\Lgen}{L_\text{gen}}
\newcommand{\Lthr}{L_{\min}}
\newcommand{\dcore}{\langle d_\text{core} \rangle}
\newcommand{\modified}[1]{{\color{red}#1}}
% \newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}

@@

%\section{Introduction}
During evolution, microbial genomes change by both local mutations and large-scale alterations \cite{arnold2021horizontal}.
Local mutations only change a few nucleotides by substitution, insertion or deletion.
Conversely, large-scale alterations reorganize the sequence, and involve either the homologous recombination of large segments, gene loss or gain, inversions, or mobilisation of genetic elements.
The accumulation of such changes over time complicates comparative genomic analyses of present day isolates.
Homologous recombination is rapid enough that most genes in many bacterial core genomes have distinct phylogenies \cite{sakoparnig2021whole} and even closely related genomes differ dramatically in gene content \cite{touchon2020phylogenetic,touchon2009organised,doolittle2009origin}.
% \cite{touchon2020phylogenetic,doolittle2009origin,touchon2009organised}.
% comment: rewritten to use consistent terminology (I think 'mutation' and 'polymorphism' for most people make them think of local changes only)
%Scalable inference techniques of this complicated relatedness structure would, in part, spur the advance of critical epidemiological surveillance tools, analogous to NextStrain, for bacterial pathogens.

@@

\section{Algorithms and implementation}
\emph{PanGraph} transforms an arbitrary set of genomes into a \emph{graph} that simultaneously compresses the collection of sequences and exhaustively summarizes both the structural and nucleotide-level polymorphisms.
The graph is composed of \emph{pancontigs}, which represent linear multiple-sequence alignments of homologous sequence found within one or more input genomes.
\emph{Pancontigs} are connected by an edge if they are adjacent in at least one input sequence; individual sequences are then recapitulated by contiguous \emph{paths} through the graph.
Pancontigs are directed, meaning their orientation in a path can be either 5' to 3' or vice versa.

To construct a pangraph, the algorithm needs to find homologous sequence within and among all input genomes.
\emph{PanGraph's} overarching strategy is to approximate multiple-genome alignment by iterative pairwise alignment of graphs of subsets of sequences, in the spirit of progressive alignment tools \cite{feng1987progressive,darling2010progressivemauve,armstrong2020progressive}.
Pairwise graph alignment is performed by an all-to-all alignment of the consensus sequence of all \emph{pancontigs} between both graphs, and the order of pairwise alignments is determined by a guide tree.

@@

\paragraph*{Merging of homologous sequence.}
If the above alignment step detected homologous stretches between the consensus sequences of two pancontigs, these pancontigs, or parts of them, can be merged.
It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.
We rank each alignment between two pancontigs according to the pseudo-energy
\begin{equation}\label{eq:pseudo-energy}
    E = -\ell + \alpha N_c + \beta N_m
\end{equation}
where $\ell$, $N_c$, and $N_m$ denote the alignment length, number of \emph{pancontigs} created by the merger, and number of polymorphisms per genome in the newly created {pancontig} respectively.
See Fig \ref{fig:visualization}B for a graphical definition for each term.
Additionally, $\alpha$ and $\beta$ are hyperparameters of the algorithm.
Only mergers whose alignment has negative pseudo-energy are performed.
The parameter $\beta$ controls the maximal sequence divergence between the two merger-candidates. The negative pseudo-energy requirement means that no merger will be performed if the Hamming distance exceeds $1/\beta$.
Importantly, this divergence threshold is applied to the comparison of the consensus sequences of the merger candidates.
The pseudo-energy ranking also effectively results in merges being performed in reverse evolutionary order, with less diverged sequences being merged first.
The parameter $\alpha$ controls the fragmentation of the graph and imposes a minimal length on the pan-contigs resulting from the merger.
The merger is always performed if it doesn't increase the number of contigs, but mergers that introduce 1, 2, 3 or 4 cuts are only performed for sufficiently long homologous stretches as parameterized by $\alpha$.
In addition to the fragmentation parameter $\alpha$, there is a further parameter $\Lthr$ that controls the minimal length of pancontigs. Structural variation at a scale smaller than this threshold (i.e. short indels) is stored in pancontig alignments, rather than in the graph structure, and no merges are performed if they have length shorter than this threshold. For instance, when a merge would generate flanking contigs shorter than this threshold, the merging is performed but the alignment is extended to include these regions. We set a default value of 100 bp for this parameter.

@@

At the graph level, the merger of two \emph{pancontigs} defines a new \emph{pancontig}, connected on both sides by edges to the neighboring \emph{pancontigs} of both inputs, and thus locally collapses the two graphs under consideration.
At the nucleotide level, the pairwise alignment of two \emph{pancontigs} maps the consensus of one onto the other; the merger of two \emph{pancontigs} requires the application of the map onto the underlying multiple-sequence alignment.
Once both sets of sequences are placed onto a common coordinate system, the resultant consensus sequence, and thus polymorphisms, are recomputed.
This procedure can be viewed as an approximate multiple sequence alignment algorithm that aligns homologous parts without further investigating insertions in the two sub-alignments. This shortens considerably the time needed, but might introduce minor inconsistencies and artifacts in the alignments. After building the graph, these can be removed by performing a multiple sequence alignment on each pancontig  (see ``polish'' command in documentation).

The above procedure is repeated until no alignments with negative energy remain.
More details on the merging procedure and the effect of parameters can be found in SI section I.

@@

\begin{figure}[tb]
    \includegraphics[width=.4\textwidth]{figs/misplaced_fraction_vs_divergence.pdf}
    \caption{{\bf Accuracy against synthetic data}.
        We generate artificial data with varying degree of sequence divergence, and compare the real underlying pangenome graph with the one reconstructed by \textit{PanGraph}, for three different alignment kernels \textit{minimap2} with \textit{asm10} or \textit{asm20} option, and \textit{mmseqs2}. In each comparison we evaluate the misplacement of breakpoints that we can pair on the two graphs within 1kbp. The plot displays the fraction of breakpoints that have misplacement greater than the standard \textit{PanGraph} precision threshold of $\Lthr=$100bp, as a function of average pairwise sequence divergence. Line and shaded area represents mean and standard deviation over 25 repetitions. \textit{mmseqs2} maintains accuracy at higher divergence, at the cost of higher computational time.
    }
    \label{fig:toy-accuracy}
\end{figure}

@@

\begin{table}[h]
    \setlength{\tabcolsep}{6pt}
    \begin{tabular}{l c c c c c c}
        \hline\hline
        Species                  & N.  & $\avg{\Lgen}$ & $\Lpang$ & $\Lcore$ & $\Lsoftcore$ & $\dcore$ \\
                                 &     & [Mbp]         & [Mbp]    & [Mbp]    & [Mbp]        &          \\
        \hline
        \textit{E.~coli}         & 307 & 5.0           & 17.7     & 0.7      & 2.0          & 1.6\%    \\
        \textit{K.~pneumoniae}   & 109 & 5.3           & 13.0     & 2.3      & 3.7          & 0.8\%    \\
        \textit{H.~pylori}       & 85  & 1.6           & 1.9      & 0.6      & 1.0          & 4.2\%    \\
        \textit{M.~tuberculosis} & 51  & 4.4           & 4.1      & 2.4      & 3.4          & 0.03\%   \\
        \textit{P.~marinus}      & 10  & 1.8           & 3.3      & 0.7      & 0.7          & 26.9\%   \\
        \hline
    \end{tabular}
    \caption{{\bf Dataset properties}. Columns represent:
        number of isolates (N.),
        average chromosome length ($\avg{\Lgen}$),
        total pangenome length ($\Lpang$),
        total core genome length ($\Lcore$),
        total soft-core genome length ($\Lsoftcore$, corresponding to genes shared by at least 95\% of the isolates),
        and average pairwise divergence of core genes ($\dcore$).
    }
    \label{table:panx-dataset}
\end{table}

@@

For species other than \textit{P.~marinus}, the pangraphs contain thousands to tens of thousands of pancontigs, and 50\% of the pangenome sequence is contained in pancontigs spanning several kbps (see SI Fig.~4).
For these species the use of null energy parameters slightly increases merging and sequence compression, at the cost of slightly more fragmented graphs (cf. SI Section IIIC and SI Fig.~4).
Overall this benchmark showcases the capabilities and limits of \textit{PanGraph}, and demonstrates the value of adapting the pseudo-energy parameter and choice of alignment kernel to the expected diversity of the dataset considered.

\begin{figure*}[htb]
    \includegraphics[width=\textwidth]{figs/panx_benchmark.pdf}
    \caption{{\bf Benchmark on real data}.
        We build pangenome graphs from fully-assembled chromosomes from 5 different bacterial species. For each species we build graphs with three different alignment kernel options (\textit{minimap2} with \textit{asm10} or \textit{asm20} option and \textit{mmseqs2}) and two different settings for the pseudo-energy parameters $\alpha, \beta$ (standard or null values).
        \textbf{A}: \textit{PanGraph} wall-time when run in parallel on 8 cores.
        \textbf{B}: fraction of core pangenome in the pangenome graph.
        \textbf{C}: sequence compression, defined as the ratio between the pangenome graph size and the cumulative size of all the sequences contained in the graph. Since maximal compression depends on the number of isolates $n$ in the pangenome graph, we mark for reference the value of $1/n$ for each dataset.
    }
    \label{fig:panx-benchmark}
\end{figure*}

@@

We then explored how the properties of the pangraph scale with the dataset size. To do so, we build pangenome graphs from randomized subsets of isolates of increasing size from the \textit{E.~coli} dataset. Results are reported in Fig.~\ref{fig:panx-size}.\\
The number and size of pancontigs scale sub-linearly with the number of isolates, with more than 50\% of the pangenome being included in the 10\% longest pancontigs. Core pancontigs having higher-than-average length. This is expected given that core genes tend to stay synteny over large evolutionary distances. Adding more strains does not generate excessive fragmentation of these pancontigs, and their size remains of several kbps even as hundreds of strains are added. The size of the core genome does not decrease significantly with the addition of new strains, while the total pangenome size remains orders of magnitude smaller than the total size of all genomes included in the graph.

@@

In addition to being a useful operation, marginalization allows us to quantify the consistency and robustness of pangraph construction on real data, where the ground truth is unknown.
To verify that marginalized graphs are compatible with newly-built graphs, we built a pangenome graph for 50 randomly selected strains from the \textit{K.~pneumoniae} dataset. We then picked 50 random pairs of strains from the same set and built two graphs for each pair: the graph obtained by marginalizing the complete graph on the pair (i.e. marginalized graph in Fig.~\ref{fig:marginalization}A top), and the graph built directly from the pair (i.e. pairwise graph in Fig.~\ref{fig:marginalization}A top).
In order to compare both graphs, we compute the partition they generated on the genome of the isolates they include. Each genome is partitioned in pancontigs that are either shared on the pair, or private to one isolate (cf. Fig.~\ref{fig:marginalization}A bottom). We can classify the segments in the intersection of the two partitions in different categories, depending on whether the two partitions agree on whether the segment is shared or private. Moreover segments on which the two partitions agree are further sub-divided depending on whether they are private or shared on both partitions.\\
The compatibility between the two graphs requires segments on which the two partitions disagree to be few and short. Indeed, we verified that over the pairs we picked, these segments cover a very small fraction of the genome ($<1\%$) and have average size compatible with the default 100 bp precision threshold of \textit{PanGraph} set by the value of $\Lthr$ (cf. Fig.~\ref{fig:marginalization}B-C). Conversely, segments on which the partitions agree have average size of several kbps. The fact that the two partitions almost completely agree cannot be explained simply by the fact that most of the genome is shared, since segments that are shared on both graphs cover on average only 85\% of the sequence. We confirmed these results using the fraction of shared k-mers, using an approach inspired by PopPUNK \cite{lees2019fast}. Namely, we approximate the fraction of homologous sequence between any two pairs using the fraction of shared k-mers ($k=21$) divided by $(1-d)^{k}$, where $d$ is the average pairwise divergence on core genes for the pair considered. This divisor corrects for k-mers on homologous sequence that are not shared due to mutations. The resulting distribution of shared sequence is in very good agreement with the fraction of segments that are shared on both graphs (cf. Fig.~\ref{fig:marginalization}B), suggesting that homologous sequences are correctly merged on the complete, marginalized and pairwise graphs.\\
We performed the same test on species from the other datasets, obtaining similar results on all species whose divergence is compatible with \textit{PanGraph} capabilities (cf. SI Section IV and SI Fig.~5).

\begin{figure*}[htb]
    \includegraphics[width=\textwidth]{figs/marginalize_test.pdf}
    \caption{{\bf Test of graph marginalization}.
        \textbf{A}: we built a pangenome graph from 50 randomly chosen strains from the \textit{K.~pneumoniae} dataset. We then randomly picked 50 pairs of strains. For each pair we compared the pangenome graph obtained by marginalizing the complete graph on the pair of strain, and the one obtained by building a new graph for the pair (top). The comparison is done by considering that each graph partitions a genome in shared and private segments. By combining the partitions generated by the marginalized and pairwise graphs we categorize segments in three categories, depending on whether the two partitions agree or not, and if they agree depending on whether segments are shared or private. All graphs were built using \textit{minimap2} alignment kernel with \textit{asm20} option and default value for the energy parameters.
        \textbf{B}: distribution of the average fraction of the genome covered by segments of each category, over the 50 pairs considered (2 entries per pair). The last line represents the distribution of shared sequence, approximated using the fraction of shared k-mers corrected using sequence divergence as described in the main text. Next to each distribution we report its mean and standard deviation.
        \textbf{C}: distribution of average segment lengths for each category over the 50 pairs considered (2 entries per pair). Mean and standard deviation are reported.
    }
    \label{fig:marginalization}
\end{figure*}

### 2023-02-24 supplementary.tex
save progress

\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\Lcore}{L_\text{core}}
\newcommand{\Lpang}{L_\text{pang}}
\newcommand{\Lgen}{L_\text{gen}}
\newcommand{\dcore}{\langle d_\text{core} \rangle}
\newcommand{\Lthr}{L_{\min}}
\newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}

@@

We used genomes from the dataset described in the previous section to verify that marginalized graphs are compatible with graphs built directly from subset of strains. For each species we selected 50 different isolates (10 for PM) and built pangenome graphs using standard parameter values and the \textit{minimap2} kernel with \textit{asm20} option. For each species we then randomly picked 50 different pairs of isolates (all 45 for PM).\footnote{The list of selected accession numbers and selected pairs for each species is available on the repository: \url{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}} For each pair we then built pairwise graphs using the same parameters, and compared them to the corresponding marginalized graph, obtained by projecting the full pangenome graph for the 50 isolates on the selected pair (cf. Fig.~6A, main text).\\
The comparison was performed by considering that each graph partitions a genome into pancontigs. Each pancontig can either be shared with the other member of the pair, or can be private to the isolate (cf. Fig.~6A, main text). For each isolate we consider the two partitions defined by the marginalized and pairwise graphs, and generate their intersection. Each block in this intersection can belong to two different categories, depending on whether the two partitions agree or disagree on whether the block is shared. Moreover, we can separate blocks on which we find agreement in two further categories: blocks that are shared and blocks that are private on both graphs.\\
For each pair of genomes and each type of block we evaluate the fraction of the genome that they cover and their average size. In Fig.~\ref{fig:marginalize-suppl} we display the distribution of these quantities across all pairs for the different species considered and the block types described above.\\
In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default pancontig length threshold of \textit{PanGraph} (\oldnew{}{$\Lthr =$}100 bp).
As described in the previous section, the divergence of the HP dataset sits on the limit of what the \textit{minimap2} alignment kernel can successfully handle. In this case the results of merging different graphs together can depend on the order of merging, and pairwise and marginalized graphs can potentially show some disagreement. In this case the disagreement remains minor and only extends to few percents of the genome size, with disagreement blocks having size only slightly bigger than the 100 bp threshold of \textit{PanGraph}.
PM represents instead an example of a dataset with divergence much beyond the limit of the alignment kernel ($\dcore \sim 27 \%$). As a consequence, highly diverged homologous sequences are not merged, and the great majority of segments are private and have large size. Results also show a strong order-dependence, with some \textit{disagreement} blocks having size of 10s of kbp.

>>>>

\newcommand{\avg}[1]{\langle #1 \rangle}
\newcommand{\Lcore}{L_\text{core}}
\newcommand{\Lpang}{L_\text{pang}}
\newcommand{\Lgen}{L_\text{gen}}
\newcommand{\dcore}{\langle d_\text{core} \rangle}
\newcommand{\Lthr}{L_{\min}}
% \newcommand{\oldnew}[2]{{\color{gray}\sout{#1}}{\color{red}#2}}

@@

We used genomes from the dataset described in the previous section to verify that marginalized graphs are compatible with graphs built directly from subset of strains. For each species we selected 50 different isolates (10 for PM) and built pangenome graphs using standard parameter values and the \textit{minimap2} kernel with \textit{asm20} option. For each species we then randomly picked 50 different pairs of isolates (all 45 for PM).\footnote{The list of selected accession numbers and selected pairs for each species is available on the repository: \url{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}} For each pair we then built pairwise graphs using the same parameters, and compared them to the corresponding marginalized graph, obtained by projecting the full pangenome graph for the 50 isolates on the selected pair (cf. Fig.~6A, main text).\\
The comparison was performed by considering that each graph partitions a genome into pancontigs. Each pancontig can either be shared with the other member of the pair, or can be private to the isolate (cf. Fig.~6A, main text). For each isolate we consider the two partitions defined by the marginalized and pairwise graphs, and generate their intersection. Each block in this intersection can belong to two different categories, depending on whether the two partitions agree or disagree on whether the block is shared. Moreover, we can separate blocks on which we find agreement in two further categories: blocks that are shared and blocks that are private on both graphs.\\
For each pair of genomes and each type of block we evaluate the fraction of the genome that they cover and their average size. In Fig.~\ref{fig:marginalize-suppl} we display the distribution of these quantities across all pairs for the different species considered and the block types described above.\\
In all species except HP and PM we find very good agreement between pairwise and marginalized graphs: segments on which the graphs \textit{disagree} represent only a very minor fraction of the genome (much smaller than the fraction of shared or private blocks), and have size compatible with the default pancontig length threshold of \textit{PanGraph} ($\Lthr =$100 bp).
As described in the previous section, the divergence of the HP dataset sits on the limit of what the \textit{minimap2} alignment kernel can successfully handle. In this case the results of merging different graphs together can depend on the order of merging, and pairwise and marginalized graphs can potentially show some disagreement. In this case the disagreement remains minor and only extends to few percents of the genome size, with disagreement blocks having size only slightly bigger than the 100 bp threshold of \textit{PanGraph}.
PM represents instead an example of a dataset with divergence much beyond the limit of the alignment kernel ($\dcore \sim 27 \%$). As a consequence, highly diverged homologous sequences are not merged, and the great majority of segments are private and have large size. Results also show a strong order-dependence, with some \textit{disagreement} blocks having size of 10s of kbp.
